An efficient composition of bidirectional programs by tupling and lazy updates

### 2019-12-02 paper/1.introduction.tex
reduce some vspaces and add some vspaces

The composition of BX programs is a fundamental construct to build more complex BX programs \cite{Bohannon06relationallenses:, Bohannon:2008:BRL:1328438.1328487}. Let $bx_1$ (defined by $get_{bx_1}$ and $put_{bx_1}$) and $bx_2$ (defined by $get_{bx_2}$ and $put_{bx_2}$) be two bidirectional programs, then their composition $bx_1 \circ bx_2$ is defined by
\begin{align}
get_{bx_1 \circ bx_2}~s &= get_{bx_2} (get_{bx_1}~s)\\
put_{bx_1 \circ bx_2}~s~v' &= put_{bx_1} s~(put_{bx_2}~ (get_{bx_1}~s)~v')
\end{align}
One feature of this composition is that $put_{bx_1 \circ bx_2}$ needs to call $get_{bx_1}$ to compute the intermediate result for $put_{bx_2}$ to use, which would introduce an efficiency problem if we compute $put$ for composition of many bidirectional programs. Generally, for a composition of $O(n)$ bidirectional programs, we need to call $get$ for $O(n^2)$ times. To be concrete, consider the evaluation of the following composition (which will be used as our running example in this paper):
\[
lp3 = (phead \circ phead) \circ phead
\]
which is illustrated by Figure \ref{fig:eval-comp-phead} with the original source $s$ being ${[[[1,2],[3]],[4]]}$ and the updated view ${100}$.
To obtain the final updated source $s'$, $put$ for $lp3$ needs to evaluate $put$ of $phead$ three times. The first is from $i_2$ and $v'$ to obtain $i'_2$, which needs to call $get$ twice to compute $i_2$; the second is from $i_1$ and $i'_2$ to obtain $i'_1$, which needs to call $get$ once, and the last is from $s$ and $i'_1$ to obtain~$s'$, which is just a direct $put$ computation.

@@

One direct solution to avoid this repeated computation of $get$ is to compute the composition in a right associative manner. For instance, if we transform  $lp3$ to $rp3$:
\[
rp3 = phead \circ (phead \circ phead)
\]
then the $put$ for $rp3$ only needs to compute $get$ of $phead$ twice, one time less than that for $lp3$.
 However, this transformation is not always easy to do. For instance, let us consider $breverse$, a bidirectional version of the traditional `reverse' program for reversing a list. It is defined using $\mbox{\it bfoldr}$, a bidirectional version of the traditional $\mbox{\it foldr}$, whose definition  is shown in the last part of Section~\ref{sect:minbigul}. Informally, $\mbox{\it bfoldr}$ is a recursive bidirectional program defined in a way like
 \[
 \mbox{\it bfoldr}~bx~\cdots = \cdots  (\mbox{\it bfoldr}~\cdots) \circ bx \cdots
 \]
where the composition is inherently left associative, and the number of composition is dynamically determined by the length of the source list. This makes it hard to do the above transformation statically. The same efficiency problem occurs in all BX languages.

>>>>

The composition of BX programs is a fundamental construct to build more complex BX programs \cite{Bohannon06relationallenses:, Bohannon:2008:BRL:1328438.1328487}. Let $bx_1$ (defined by $get_{bx_1}$ and $put_{bx_1}$) and $bx_2$ (defined by $get_{bx_2}$ and $put_{bx_2}$) be two bidirectional programs, then their composition $bx_1 \circ bx_2$ is defined by

\minusvspace
\begin{align}
get_{bx_1 \circ bx_2}~s &= get_{bx_2} (get_{bx_1}~s)\\
put_{bx_1 \circ bx_2}~s~v' &= put_{bx_1} s~(put_{bx_2}~ (get_{bx_1}~s)~v')
\end{align}
\minusvspace

\noindent One feature of this composition is that $put_{bx_1 \circ bx_2}$ needs to call $get_{bx_1}$ to compute the intermediate result for $put_{bx_2}$ to use, which would introduce an efficiency problem if we compute $put$ for composition of many bidirectional programs. Generally, for a composition of $O(n)$ bidirectional programs, we need to call $get$ for $O(n^2)$ times. To be concrete, consider the evaluation of the following composition (which will be used as our running example in this paper):

\minusvspacetwo
\[
lp3 = (phead \circ phead) \circ phead
\]
\minusvspacetwo

\noindent which is illustrated by Figure \ref{fig:eval-comp-phead} with the original source $s$ being ${[[[1,2],[3]],[4]]}$ and the updated view ${100}$.
To obtain the final updated source $s'$, $put$ for $lp3$ needs to evaluate $put$ of $phead$ three times. The first is from $i_2$ and $v'$ to obtain $i'_2$, which needs to call $get$ twice to compute $i_2$; the second is from $i_1$ and $i'_2$ to obtain $i'_1$, which needs to call $get$ once, and the last is from $s$ and $i'_1$ to obtain~$s'$, which is just a direct $put$ computation.

@@

One direct solution to avoid this repeated computation of $get$ is to compute the composition in a right associative manner. For instance, if we transform  $lp3$ to $rp3$:

\minusvspacetwo
\vspace{-4mm}
\[
rp3 = phead \circ (phead \circ phead)
\]
\minusvspacetwo

\noindent then the $put$ for $rp3$ only needs to compute $get$ of $phead$ twice, one time less than that for $lp3$.
However, this transformation is not always easy to do. For instance, let us consider $breverse$, a bidirectional version of the traditional `reverse' program for reversing a list. It is defined using $\mbox{\it bfoldr}$, a bidirectional version of the traditional $\mbox{\it foldr}$, whose definition  is shown in the last part of Section~\ref{sect:minbigul}. Informally, $\mbox{\it bfoldr}$ is a recursive bidirectional program defined in a way like

\minusvspacetwo
 \[
 \mbox{\it bfoldr}~bx~\cdots = \cdots  (\mbox{\it bfoldr}~\cdots) \circ bx \cdots
 \]
 \minusvspacetwo

\noindent where the composition is inherently left associative, and the number of composition is dynamically determined by the length of the source list. This makes it hard to do the above transformation statically. The same efficiency problem occurs in all BX languages.

### 2019-12-02 paper/2.minbigul.tex
reduce some vspaces and add some vspaces

    $\qtab \getbxinline{bx} : s \to v, \quad \putbxinline{bx} : s * v \to v$

@@

    $\qtab \putbx{bx}{s}{(\getbx{bx}{s})} = s \qquad [\textsc{GetPut}]$

    $\qtab \getbx{bx}{(\putbx{bx}{s}{v})} = v \qquad [\textsc{PutGet}]$

@@

    $\qtab \putbx{bx}{(\putbx{bx}{s}{v'})}{v} = \putbx{bx}{s}{v} \qquad [\textsc{PutPut}]$

@@

    $bx := Skip \ h \
        | \ Replace \
        | \ Prod \ bx_1 \ bx_2 \
        | \ RearrS \ f_1 \ f_2 \ bx \
        | \ RearrV \ g_1 \ g_2 \ bx \\
    \qtabs{2} | \ Case \ cond_{sv} \ cond_{s} \ bx_1 \ bx_2 \
        | \ Compose \ bx_1 \ bx_2$

@@

$\tab \text{if } E_1 \text{ then } X_1 \text{ else } X_2 \text{ fi } E_2; S\\
\tabs{2} \ \equiv \text{if }E_1 = true \text{ then } \{ X_1; \text{ if } E_2 = true \text{ then } S \text{ else assert } false \}\\
    \tabs{4} \text{else } \{ X_2; \text{ if } E_2 = false \text{ then } S \text{ else assert } false \}$

\noindent As an example of minBiGUL program, consider the definition of $phead$ :\\
$\tab phead = RearrS \ f_1 \ f_2 \ bx_s \ \text{where: } f_1 = \lambda (s::ss).(s,ss), \, f_2 = \lambda (s,ss).(s::ss),\\
    \qtab bx_s = RearrV \ g_1 \ g_2 \ bx_v \ \text{where: } g_1 = \lambda v.(v,()), \, g_2 = \lambda (v,()).v,\\
        \qtabs{2} bx_v = \product{Replace}{(Skip \ (\lambda \_.())}$

@@

$\tab \putbx{phead \circ phead}{[[1,2,3],[\,],[4,5]]}{100} = [[100,2,3],[\,],[4,5]]$

$\tab \getbx{phead \circ phead}{[[1,2,3],[\,],[4,5]]} = 1$

@@

$\tab bfoldr \ bx = Case \ cond_{sv} \ cond_s \ bx_1 \ bx_2$ where:\\
$\tabs{3} \ cond_{sv} = \lambda (s_1,s_2).\lambda v.(s_1 = [\ ]), \ cond_s = \lambda (s_1,s_2).(s_1 = [\ ])$\\
$\tabs{3} \ bx_1 = RearrV \ g_1 \ g_2 \ bx_v$ where: \\
    $\tabs{4} g_1 = g_2^{-1} = \lambda [v].(v,[\ ]), bx_v = (Skip \ (\lambda \_.())) \times Replace$\\
$\tabs{3} \ bx_2 = RearrS \ f_1 \ f_2 \ bx_s$ where:\\
    $\tabs{4} f_1 = f_2^{-1} = \lambda ((x:xs),e).(x,(xs,e))$, 
    $bx_s = ((Replace \times bfoldr \ bx) \circ bx)$

>>>>

 \smallvspace
    $\qtab \getbxinline{bx} : s \to v, \quad \putbxinline{bx} : s * v \to v$
 \smallvspace

@@

\smallvspace
    $\qtab \putbx{bx}{s}{(\getbx{bx}{s})} = s \qquad [\textsc{GetPut}]$

    $\qtab \getbx{bx}{(\putbx{bx}{s}{v})} = v \qquad [\textsc{PutGet}]$
\smallvspace

@@

\smallvspace
$\qtab \putbx{bx}{(\putbx{bx}{s}{v'})}{v} = \putbx{bx}{s}{v} \qquad [\textsc{PutPut}]$
\smallvspace

@@

\smallvspace
    $bx := Skip \ h \
        | \ Replace \
        | \ Prod \ bx_1 \ bx_2 \
        | \ RearrS \ f_1 \ f_2 \ bx \
        | \ RearrV \ g_1 \ g_2 \ bx \\
    \qtabs{2} | \ Case \ cond_{sv} \ cond_{s} \ bx_1 \ bx_2 \
        | \ Compose \ bx_1 \ bx_2$
\smallvspace

@@

\smallvspace
$\tab \text{if } E_1 \text{ then } X_1 \text{ else } X_2 \text{ fi } E_2; S\\
\tabs{2} \ \equiv \text{if }E_1 = true \text{ then } \{ X_1; \text{ if } E_2 = true \text{ then } S \text{ else } undefined \}\\
    \tabs{4} \text{else } \{ X_2; \text{ if } E_2 = false \text{ then } S \text{ else } undefined \}$
\smallvspace

\noindent As an example of minBiGUL program, consider the definition of $phead$ :\\
\smallvspace
$\tab phead = RearrS \ f_1 \ f_2 \ bx_s \ \text{where: } f_1 = \lambda (s::ss).(s,ss), \, f_2 = \lambda (s,ss).(s::ss),\\
    \qtab bx_s = RearrV \ g_1 \ g_2 \ bx_v \ \text{where: } g_1 = \lambda v.(v,()), \, g_2 = \lambda (v,()).v,\\
        \qtabs{2} bx_v = \product{Replace}{(Skip \ (\lambda \_.())}$
\smallvspace

@@

\smallvspace
$\tab \putbx{phead \circ phead}{[[1,2,3],[\,],[4,5]]}{100} = [[100,2,3],[\,],[4,5]]$

$\tab \getbx{phead \circ phead}{[[1,2,3],[\,],[4,5]]} = 1$
\smallvspace

@@

\smallvspace
$\tab bfoldr \ bx = Case \ cond_{sv} \ cond_s \ bx_1 \ bx_2$ where:\\
$\tabs{3} \ cond_{sv} = \lambda (s_1,s_2).\lambda v.(s_1 = [\ ]), \ cond_s = \lambda (s_1,s_2).(s_1 = [\ ])$\\
$\tabs{3} \ bx_1 = RearrV \ g_1 \ g_2 \ bx_v$ where: \\
    $\tabs{4} g_1 = g_2^{-1} = \lambda [v].(v,[\ ]), bx_v = (Skip \ (\lambda \_.())) \times Replace$\\
$\tabs{3} \ bx_2 = RearrS \ f_1 \ f_2 \ bx_s$ where:\\
    $\tabs{4} f_1 = f_2^{-1} = \lambda ((x:xs),e).(x,(xs,e))$, 
$bx_s = ((Replace \times bfoldr \ bx) \circ bx)$
\smallvspace
